Space Matters: extending sensitivity analysis to initial spatial conditions in geosimulation models

### 2018-09-28 Response_to_Reviewers.tex

   % I believe it would be better to begin with an illustrative example on how two or three initial patterns influence the model results and then use the suggested method for a more extensive analysis
   \item We agree that illustrative examples would provide a relevant starting point to  the analysis but we were afraid that their addition to the paper would make the introduction less straightforward and have the opposite effect to the one we were seeking. As we were nevertheless very interested by your remark, we provide a short answer to your proposition.

In order to get a first idea of the potential impact of spatial initial conditions on to the results of the simulation, we implemented on Segregation model a geographical constraint: some cells cannot be inhabited (yellow cells in Figure \ref{Fig1}). The way those “forbidden” cells are distributed throughout space will indeed influence the steps of the model. 
We introduced various spatial initial conditions in the model that we controlled by one clustering parameter on the forbidden cells. A value of 0 would indicate a spatial distribution of yellow cells so as to minimize spatial autocorrelation. A value of 1 indicates on the contrary a concentration process that tends towards a high value of autocorrelation (see below Figure \ref{Fig1}).

We then compute the number of communities arising when the model converge, hence the number of different related components of similar neighbors.
Using 270 replications of 11 values of the clustering parameter (from 0 to 1 by step 0.1), we obtained a significant growth in terms of number of communities when the constraints on residential locations are concentrated in space, which might be interpreted in such example as a bigger segregation when density of housing is heterogeneously distributed through space. 
%In the rest of the article, we wish to further investigate such spatial effects in a much more systematic way.

%%%%%%%%%%%%%%
\begin{minipage}{\textwidth}
\begin{center} 
\resizebox{0.9\textwidth}{!}{ 
	\includegraphics{figures/SpaceMatters2.bmp}
} \textbf{Effects of clustering of forbidden cells on the number of communities emerging from an adapted Schelling model (implementation on Netlogo).} \label{fig:Fig1}
\end{center}
\end{minipage}
%%%%%%%%%%%%%%

	% 2. It is very important to describe the models. This is especially relevant for the Schelling model, which has many variants; each has different dependencies on the initial pattern. In my opinion, the NetLogo implementation is not the best choice because the model does not reach a state of segregation when agents are highly intolerant.
	\item There was a misunderstanding for the Schelling implementation actually used, as we used our own implementation which pseudo-code has been added in Appendix of the paper and source code is openly available on the git repository of the project. This version implements the one used by (Gauvin et al., 2009), which also influenced us for the use of the term ``phase diagram''.

### 2018-09-28 Response_to_Reviewers.tex

Concerning specifically the comments made by the first referee:

\begin{itemize}
	% 1. I think it would be better if the author would focus on a single model. This way they could describe the model and the analyses in more detail. As it stands, the models are treated as black boxes. 
	\item The suggestion to focus on a single model is a qualified comment. However, after presenting our research to several audiences, we came to the conclusion that the genericity of our approach would be better served by two case studies rather than with a single model. In that perspective, the ``black-box'' is seen as an advantage as it reinforces genericity. We acknowledge the fact that this choice led us to shorten the model presentations. Nevertheless, we have taken notice of your remark by improving the description of the models and their results through the addition of pseudo-codes in Appendix, result descriptions in the text and by recalling more clearly the link to the model source we used. Furthermore, we have focused on some of the most classical models of the ABM literature so as to lower the reader's cost of entry into the methodology.

   % I believe it would be better to begin with an illustrative example on how two or three initial patterns influence the model results and then use the suggested method for a more extensive analysis
   \item We agree that illustrative examples would provide a relevant starting point to  the analysis but we were afraid that their addition to the paper would make the introduction less straightforward and have the opposite effect to the one we were seeking. As we were nevertheless very interested by your remark, we provide a short answer to your proposition.

In order to get a first idea of the potential impact of spatial initial conditions on to the results of the simulation, we implemented on Segregation model a geographical constraint: some cells cannot be inhabited (yellow cells in Figure 1). The way those “forbidden” cells are distributed throughout space will indeed influence the steps of the model. 
We introduced various spatial initial conditions in the model that we controlled by one clustering parameter on the forbidden cells. A value of 0 would indicate a spatial distribution of yellow cells so as to minimize spatial autocorrelation. A value of 1 indicates on the contrary a concentration process that tends towards a high value of autocorrelation (see below Figure 1). % note : no fig envt in moderncv

@@

We then compute the number of communities arising when the model converge, hence the number of different related components of similar neighbors.
Using 270 replications of 11 values of the clustering parameter (from 0 to 1 by step 0.1), we obtained a significant growth in terms of number of communities when the constraints on residential locations are concentrated in space, which might be interpreted in such example as a bigger segregation when density of housing is heterogeneously distributed through space. 
%In the rest of the article, we wish to further investigate such spatial effects in a much more systematic way.

%%%%%%%%%%%%%%
\begin{minipage}{\textwidth}
\begin{center} 
	\includegraphics[width=\textwidth]{figures/SpaceMatters2.png}

\textbf{Effects of clustering of forbidden cells on the number of communities emerging from an adapted Schelling model (Netlogo implementation).}
\end{center}
\end{minipage}
%%%%%%%%%%%%%%

@@

	% Last paragraph of page 5, you do not state explicitly if genetic algorithms were use within the OpenMOLE platform.
	\item The functionalities of OpenMOLE we used were detailed in the article; we did not use genetic algorithms but sampling methods and high performance computing environments.

	% In section 3.1, page 8, line 404-405 the authors mention that the distance of the phase diagrams was measured with respect to the reference phase diagram computed on the default initial spatial conditions. Actually this came as a surprise to me. I had the impression that all phase diagrams would be evaluated against all other diagrams. Perhaps this should be explained earlier in the paper.
	\item \comment[JR]{ce qu'il suggère c'est un indice intégré sur l'ensemble des phases diagramme, donc un ``meta-meta-indice'' en quelque sorte, nous on reste à un niveau de meta seulement car on regarde juste pour chaque s'il bouge par rapport à un point de reference. En termes thematique c'est plus coherent pour nous, lui doit avoir en tete des indices de sensibilite globaux type Saltelli. $\rightarrow$ a voir comment justifier ca bien}{(JR) inclure en mise en contexte ce lien avec les Saltelli et indexes globaux}\comment[FL]{oui je trouve qu'on a le droit de rester limiter à cela c'est déjà pas mal!}

	% Figure 4 and 5: I find it confusing that the green color indicates higher distance values and the red color indicates lower values. I think this should be reversed.
    \item The double color progression in Fig 4 and 5 is used to highlight the opposition between ratio values above/under 1 but is not meant to introduce any hierarchy in the interpretation. We reversed the order of color according to your remark.
    %We would therefore prefer to keep the figures as they are if this is ok for you?

### 2018-09-28 Response_to_Reviewers.tex

   % I believe it would be better to begin with an illustrative example on how two or three initial patterns influence the model results and then use the suggested method for a more extensive analysis
   \item We agree that illustrative examples would provide a relevant starting point to  the analysis but we were afraid that their addition to the paper would make the introduction less straightforward and have the opposite effect to the one we were seeking. As we were nevertheless very interested by your remark, we provide a short answer to your proposition.

In order to get a first idea of the potential impact of spatial initial conditions on to the results of the simulation, we implemented on Segregation model a geographical constraint: some cells cannot be inhabited (yellow cells in Figure 1). The way those “forbidden” cells are distributed throughout space will indeed influence the steps of the model. 
We introduced various spatial initial conditions in the model that we controlled by one clustering parameter on the forbidden cells. A value of 0 would indicate a spatial distribution of yellow cells so as to minimize spatial autocorrelation. A value of 1 indicates on the contrary a concentration process that tends towards a high value of autocorrelation (see below Figure 1). % note : no fig envt in moderncv

We then compute the number of communities arising when the model converge, hence the number of different related components of similar neighbors.
Using 270 replications of 11 values of the clustering parameter (from 0 to 1 by step 0.1), we obtained a significant growth in terms of number of communities when the constraints on residential locations are concentrated in space, which might be interpreted in such example as a bigger segregation when density of housing is heterogeneously distributed through space. 
%In the rest of the article, we wish to further investigate such spatial effects in a much more systematic way.

%%%%%%%%%%%%%%
\begin{minipage}{\textwidth}
\bigskip\bigskip
\begin{center} 
	\includegraphics[width=\textwidth]{figures/SpaceMatters2.png}

Fig. 1: \textbf{Effects of clustering of forbidden cells on the number of communities emerging from an adapted Schelling model (Netlogo implementation).}
\end{center}
\bigskip\bigskip
\end{minipage}
%%%%%%%%%%%%%%

@@

	% The authors do not differentiate between data driven empirical models versus theory-driven models. I think a clear differentiation between the two groups and the positioning of the two case studies in the second group would make the paper more balanced.
	\item Concerning the remarks on synthetic population generation and the different types of models, we clearly stated our positioning regarding models built from theory, for which the proposed sensitivity analysis workflow is more suited and that our two case studies illustrate.

	% In section 1.3 (Previous considerations on the effects of the spatial configuration in simulation models) authors provide a lot of evidence on previous work on the impact of initial spatial conditions on the output of models. This is contrary to the statement in the abstract that initial conditions are often taken for granted.
	\item Indeed, we have tried to review articles where some kind of sensitivity to initial spatial conditions are tackled. However we would like to emphasize that the references cited in this section are few compared to the mass of geosimulation articles published with hardly any mention of dependence of output on initial spatial condition. Moreover, even if such sensitivity is analysed, it mostly provide some hints of impact of initial spatial conditions on the output of models (due to the accuracy of data, shape of boundaries and spatial heterogeneity) but to the best of our knowledge no article provide a systematic inquiry into the impact of systematically defined spatial input on the model output in the way we try to provide later on in the paper. We have tried to argue this point at the end of the section.

	% The most interesting part of the research to me, is not the way the density layers are generated but the use of the phase diagrams, defined as a vector of final aggregated model outputs. As these are vectors, they are not so different from time series and many techniques exist to compare vectors in this domain. I wonder if the Euclidean distance is the best measure to determine the difference between the phase vectors of different spatial inputs.
	\item The comment on other possible techniques to compare phase diagrams echoes one other reviewer comment, and we indeed expanded the discussion on this point, showing first that other distances do not change much qualitative results (analysis added in supplementary material) and with more literature for possible extensions.

@@

\begin{itemize}
	% In sensitivity analysis of spatial models, stochastic effects and small parameter changes are normally applied. What about initial conditions? The authors test the effects of different arrangements (e.g. monocentric vs. polycentric density patterns) on agent-based Schelling segregation models and Sugarscape's unequal societies model. Overall, I find this to be a very good paper. Very well written in general. Good review of the problem, clearly outlining (with references) the need for focus on initial conditions in geosimulation. The chief factors (data uncertainty, nature of spatial units (e.g. shape) and spatial heterogeneity) on initial spatial conditions are covered: -how many of these do you test? I see heterogeneity addressed, but what about shape and uncertainty? You touch upon shape when talking about NetLogo's use of raster as opposed to vector in the Discussion - could you expand on this?
	\item We thank the reviewer for this good remark. We provide references to papers addressing these issues. However, we have not addressed the shape of cells and the uncertainty of data in this paper, only the diversity of shapes of urban morphologies (e.g. heterogeneity).

	% Rigorous experimental method described. In the matching of European metropolitan areas to your generated outputs, I wonder what you are missing by not using cities from other continents. i.e. by widening the cities available to your study it may be possible to even better discover real-world examples of your archtypes (compact, polycentric, discontinuous).
	\item To the best of our knowledge, there is no study of urban morphologies similar to the one we used but at a global scale (this should be linked to the absence of a raster of population with a resolution fine enough). Working on other regional systems of cities such as the US could be possible but remains out of the scope of this paper. We added this interesting suggestion to the discussion.

	% Why do you not have the same amount of repetitions for your Sugarscape (50 repetitions) and Schelling (100 repetitions) model runs?
	\item Concerning the number of repetitions, statistical analysis show that Schelling indicators need more repetitions for an equivalent size of confidence interval on the estimated mean, as the Sharpe ratios are mostly lower than for Sugarscape. This remark was added in the paper and the statistical analysis as a supplementary material.

	% You say your models are open source - do you give details of that source?  
	\item Concerning the source code of the models, an open public repository was put into place and the link provided in the paper (to an anonymized version to ensure blind review). Similarly, the large simulation datasets used for results presented in the paper are openly available on a dataverse repository (provisorily anonymized also).

### 2018-09-28 paper.tex
.

Finally, although this may seem obvious, cities are not regular grids, and the distribution of density (of jobs, residents, buildings, etc.) is far from isotropic, even in \deleted{the most }sprawled cities. On the contrary, there is a significant diversity in the way people, activities and structures are distributed in cities. In Europe for example, \citet{LeNechet2015} quantifies and classifies six broad types of residential density distributions. However, most geographical models, especially cellular automata, still represent cities as uniform grids of isotropic density. Even in applied cases when GIS geometries of a particular city are used, the spatial distribution of agents \replaced{tend to be}{is still} approximated by a constant density \citep{arribas2014diverse}, although previous research shows that it is computationally and methodologically feasible to use accurate locations in a simple model such as Schelling's \citep{benenson2002entity}. The isotropic simplification is potentially harmful to the representation of urban processes \deleted{under inquiry }because density and accessibility have environmental, economic and social consequences. Additionally, we expect the initial spatial distribution of agents to influence simulation results in the long run \citep{Castellanoetal2009}, because the agents' rule of action itself may depend on the spatial structure of the environment. For example, households can have different preferences with respect to the built-environment they \added{might} want to live in \citep{SpielmanHarrison2014}, or agents \replaced{moving around will sense a different set of objects within the same fixed radius depending on}{can move around and sense their environment up to a certain distance, so that the resulting number of sensed objects will depend on} the topology \citep{Banos2012} and distribution of density of the sensed environment \citep{LauriJaggi2003, FossettDietrich2009}. The way modellers represent \added{initial} space is therefore a central element of geographical simulation models. However, this step is rarely explicit. \deleted{For example

and authors tend to generalise the conclusion of the model from a single spatial case study (be it a theoretical pattern or a singular empirical one. }A meaningful way of representing \added{initial} space might \deleted{thus }be to consider, not necessarily the peculiarities of every city, but at least their broad density structures so as to estimate the variability of the model behaviour \replaced{to different}{in} plausible spatial arrangements.

@@

\newpage

@@

For both models we estimate the Sharpe ratios for each indicator by $S(X) = \hat{\Eb{X}}/\hat{\sigma}(X)$ with standard estimators for average and standard deviation. The summary statistics computed on repetitions for all parameter points are given in Table~\ref{tab:summarysharpes}.

%%%%%%%%%%%%%%
\begin{table*}
\caption{\added{\textbf{Summary statistics of Sharpe ratio estimated on repetitions for each parameter point.}}\label{tab:summarysharpes}}
	\begin{tabular}{|cccccc|}
Model/Indicator & Min. & 1st Qu. &  Median &  Mean & 3rd Qu. & Max. \\\hline
Sugarscape/Gini & 2.343 & 14.413 & 20.993 & 20.098 & 26.208 & 49.808\\\hline 
Schelling/Dissimilarity & 1.053 & 12.368 & 19.434 & 25.268 & 24.328 & 468.563\\
Schelling/Entropy & 0.5458 & 8.2591 & 14.5112 & 17.0100 & 17.9372 & 244.3230\\
Schelling/Moran & 0.0001 & 0.5554 & 0.7782 & 3.2724 & 3.8979 & 121.6829\\\hline
    \end{tabular}
\end{table*}
%%%%%%%%%%%%%%

@@

\begin{table*}
	\caption{\added{\textbf{Summary statistics of different distances for schelling.}}}
	\begin{tabular}{|ccccccc|}
    	$q$ & Min. & 1st Qu. & Median & Mean & 3rd Qu. & Max. \\\hline
        1 & 0.00000 & 0.02838 & 0.06263 & 0.10608 & 0.16818 & 0.37018 \\
        2 & 0.0000 & 0.1520 & 0.2517 & 0.3107 & 0.4359 & 0.8155 \\
        3 & 0.0000 & 0.3709 & 0.5133 & 0.5860 & 0.7435 & 1.2930 \\
        10 & 0.000 & 2.083 & 2.431 & 2.380 & 2.713 & 3.664 \\\hline
    \end{tabular}
\end{table*}

\begin{table*}
	\caption{\added{\textbf{Summary statistics of different distances for sugarscape.}}}
	\begin{tabular}{|ccccccc|}
    	$q$ & Min. & 1st Qu. & Median & Mean & 3rd Qu. & Max. \\\hline
        1 & 0.07184 & 0.13996 & 0.39816 & 0.42452 & 0.64751 & 1.13169\\
        2 & 0.08909 & 0.19794 & 1.52272 & 1.29594 & 2.16371 & 2.98273\\
        3 & 0.1025 & 0.2288 & 3.1495 & 2.5074 & 4.2718 & 5.5378\\
        10 & 0.1540 & 0.4466 & 9.8757 & 7.6246 & 13.0550 & 16.8001\\\hline
    \end{tabular}
\end{table*}

@@

%%%%%%%%%%%%%%%
\begin{figure*}
	\caption{\added{\textbf{Pseudo-code (NetLogo style) for the Schelling model used here.}}}
	\begin{mdframed}
	\begin{lstlisting}
	while (*@\textit{not stopping-condition}@*) [
    	ask turtles [
    		set unsatisfied? false
    		let neighboragents turtles in-radius neighborhood-radius
    		let wantedcolor color
    		let stranger-rate count neighboragents with [color = wantedcolor] / count neighboragents
    		if stranger-rate < similar-wanted [
    set unsatisfied? true random-move
    ]
  ]
    \end{lstlisting}
    \end{mdframed}
\end{figure*}
%%%%%%%%%%%%%%%

%%%%%%%%%%%%%%%
\begin{figure*}
% Minimal netlogo code for the sugarscape model
\caption{\added{\textbf{Pseudo-code (NetLogo style) for the sugarscape model.} We give a pseudo-code very close to the NetLogo language, at the exception of italic expressions which are not put here in NetLogo to ease understandability.}}
\begin{mdframed}
\begin{lstlisting}
while [time < final-time] [
	; regrow the patch resources
    ask patches [set psugar min (list max-psugar (psugar + 1))]

>>>>

Finally, although this may seem obvious, cities are not regular grids, and the distribution of density (of jobs, residents, buildings, etc.) is far from isotropic, even in \deleted{the most }sprawled cities. On the contrary, there is a significant diversity in the way people, activities and structures are distributed in cities. In Europe for example, \citet{LeNechet2015} quantifies and classifies six broad types of residential density distributions. However, most geographical models, especially cellular automata, still represent cities as uniform grids of isotropic density. Even in applied cases when GIS geometries of a particular city are used, the spatial distribution of agents \replaced{tend to be}{is still} approximated by a constant density \citep{arribas2014diverse}, although previous research shows that it is computationally and methodologically feasible to use accurate locations in a simple model such as Schelling's \citep{benenson2002entity}. The isotropic simplification is potentially harmful to the representation of urban processes \deleted{under inquiry }because density and accessibility have environmental, economic and social consequences. Additionally, we expect the initial spatial distribution of agents to influence simulation results in the long run \citep{Castellanoetal2009}, because the agents' rule of action itself may depend on the spatial structure of the environment. For example, households can have different preferences with respect to the built-environment they \added{might} want to live in \citep{SpielmanHarrison2014}, or agents \replaced{moving around will sense a different set of objects within the same fixed radius depending on}{can move around and sense their environment up to a certain distance, so that the resulting number of sensed objects will depend on} the topology \citep{Banos2012} and distribution of density of the sensed environment \citep{LauriJaggi2003, FossettDietrich2009}. The way modellers represent \added{initial} space is therefore a central element of geographical simulation models. However, this step is rarely explicit. \deleted{For example and authors tend to generalise the conclusion of the model from a single spatial case study (be it a theoretical pattern or a singular empirical one. }A meaningful way of representing \added{initial} space might \deleted{thus }be to consider, not necessarily the peculiarities of every city, but at least their broad density structures so as to estimate the variability of the model behaviour \replaced{to different}{in} plausible spatial arrangements.

@@

%\newpage

@@

\added{For both models we estimate the Sharpe ratios for each indicator by $S(X) = \hat{\mathbb{E}{X}}/\hat{\sigma}(X)$ with standard estimators for average and standard deviation. The summary statistics computed on repetitions for all parameter points are given in Table~\ref{tab:summarysharpes}.}

%%%%%%%%%%%%%%
\begin{table*}
\caption{\added{\textbf{Summary statistics of Sharpe ratio estimated on repetitions for each parameter point.}}\label{tab:summarysharpes}}
\centering
	\begin{tabular}{|ccccccc|}
	\hline
Model/Indicator & Min. & 1st Qu. &  Median &  Mean & 3rd Qu. & Max. \\\hline
Sugarscape/Gini & 2.343 & 14.413 & 20.993 & 20.098 & 26.208 & 49.808\\\hline 
Schelling/Dissimilarity & 1.053 & 12.368 & 19.434 & 25.268 & 24.328 & 468.563\\
Schelling/Entropy & 0.5458 & 8.2591 & 14.5112 & 17.0100 & 17.9372 & 244.3230\\
Schelling/Moran & 0.0001 & 0.5554 & 0.7782 & 3.2724 & 3.8979 & 121.6829\\\hline
    \end{tabular}
\end{table*}
%%%%%%%%%%%%%%

@@

\begin{table*}
	\caption{\added{\textbf{Summary statistics of different distances for schelling.}}}
	\centering
	\begin{tabular}{|ccccccc|}
	\hline
    	$q$ & Min. & 1st Qu. & Median & Mean & 3rd Qu. & Max. \\\hline
        1 & 0.00000 & 0.02838 & 0.06263 & 0.10608 & 0.16818 & 0.37018 \\
        2 & 0.0000 & 0.1520 & 0.2517 & 0.3107 & 0.4359 & 0.8155 \\
        3 & 0.0000 & 0.3709 & 0.5133 & 0.5860 & 0.7435 & 1.2930 \\
        10 & 0.000 & 2.083 & 2.431 & 2.380 & 2.713 & 3.664 \\\hline
    \end{tabular}
\end{table*}

\begin{table*}
	\caption{\added{\textbf{Summary statistics of different distances for sugarscape.}}}
	\centering
	\begin{tabular}{|ccccccc|}
	\hline
    	$q$ & Min. & 1st Qu. & Median & Mean & 3rd Qu. & Max. \\\hline
        1 & 0.07184 & 0.13996 & 0.39816 & 0.42452 & 0.64751 & 1.13169\\
        2 & 0.08909 & 0.19794 & 1.52272 & 1.29594 & 2.16371 & 2.98273\\
        3 & 0.1025 & 0.2288 & 3.1495 & 2.5074 & 4.2718 & 5.5378\\
        10 & 0.1540 & 0.4466 & 9.8757 & 7.6246 & 13.0550 & 16.8001\\\hline
    \end{tabular}
\end{table*}

@@

%%%%%%%%%%%%%%%
\begin{figure*}
\caption{\added{\textbf{Pseudo-code (NetLogo style) for the Schelling model used here.} We give a pseudo-code very close to the NetLogo language, at the exception of italic expressions which are not put here in NetLogo to ease understandability.}}
\begin{mdframed}
\begin{lstlisting}
while (*@\textit{not stopping-condition}@*) [
	ask turtles [
		set unsatisfied? false
		let neighboragents turtles in-radius neighborhood-radius
		let wantedcolor color
		let stranger-rate count neighboragents with [color = wantedcolor] /
			count neighboragents
		if stranger-rate < similar-wanted [
			set unsatisfied? true random-move
		]
	]
    \end{lstlisting}
    \end{mdframed}
\end{figure*}
%%%%%%%%%%%%%%%

%%%%%%%%%%%%%%%
\begin{figure*}
% Minimal netlogo code for the sugarscape model
\caption{\added{\textbf{Pseudo-code (NetLogo style) for the sugarscape model.} Conventions are the same than for Schelling pseudo-code.}}
\begin{mdframed}
\begin{lstlisting}
while [time < final-time] [
;regrow the patch resources
    ask patches [set psugar min (list max-psugar (psugar + 1))]
